---
title: "EDSD 2024–2025: Sequence Analysis' Assignment"
author:
  - "Thiago C. Almeida"
date: "`r Sys.Date()`"
format:
  pdf:
    pdf-engine: xelatex
    number-sections: true
    toc: true
    header-includes:
      - \KOMAoption{captions}{tableheading}
      - |
        \usepackage{fancyhdr}
        \usepackage{graphicx}
        \usepackage{fvextra}
        \pagestyle{fancy}
        \fancyhf{}
        \fancyhead[L]{Sequence Analysis' Assignment}
        \fancyhead[R]{\includegraphics[height=30pt]{edsd_logo.jpeg}}
        \fancyfoot[C]{\thepage}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
          commandchars=\\\{\},
          breaklines=true,
          breakanywhere=true,
          fontsize=\small
        }
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# setups
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r packages, include=FALSE}

# general settings
rm(list = ls()) # removing all items of environment
invisible(gc()) # cleaning the storage memory of the computer

# packages
library(pacman)
p_load(tidyverse, TraMineR, RColorBrewer, WeightedCluster, cluster, haven, Hmisc, ggdendro, patchwork)

# project roots
project_root <- here::here()
setwd(project_root)

# general setups for graphs
graph_size = 8

```

```{r load-data, include=FALSE}
load("outputs_sequence.RData")
df <- read_dta(
  "../../../data/PartnerBirthbio.dta"
)
```

For this assignment, we will work with data provided to illustrate life course trajectories of a sample of individuals from the German Family Panel (pairfam) data.

We will work with four different partnership states to construct the trajectories:

  - $S$: Single
  
  - $LAT$: Living Apart Together
  
  - $COH$: Cohabiting
  
  - $MAR$: Married

```{r data-wrangling, include=FALSE}


# reconding variable

df <- df |>
  mutate_all(
    ~(case_when(
      . < 3 ~ 1,            # Single
      . %in% c(3,4) ~ 2,    # LAT
      . %in% c(5,6) ~ 3,    # Cohabiting
      . > 6 ~ 4,           # Married
      TRUE ~ NA_integer_)
  )) |>
  na.omit()


```

Once we identified some missing values either for the states or for individual weights, we opted to exclude these individuais of our sample.

We ended up with `r paste0(round(dim(df)[1],0))` individuais to compute the sequences.

# 1. Import the data in R and describe the distribution of life states using relevant sequence analysis tools, including tables, indexes and figures.

First step is to create the dataset based on the sequences to describe the distribution of life states.

```{r}
sa.states <- c("S", "LAT", "COH", "MAR")
sa.labels <-  c("Single", "LAT", "Cohabiting", "Married")

# create state sequence object
df.seq <- seqdef(
  df,
  labels = sa.labels,
  states = sa.states,
  weights = df$weight40
)
```

To describe the distribution of life course states, we can use different measures.

  1. Time spent in different states

```{r}

# 1 - Time spent in different states

seqmeant(seqdss(df.seq),serr = TRUE)
```

  2. Number of transitions

```{r}
# 2 - Number of transitions

wtd.mean(seqtransn(df.seq), df$weight40)

```

  3. Transition matrix between states

```{r}
# 3 - Transition matrix between states

seqtrate(df.seq, weighted = TRUE)

```

  4. State distribution at different ages (measured in months since the individuais started to be followed)

```{r}
# 4 - State distribution at different ages

seqstatd(df.seq)$Frequencies[,c(1, seq(24, 264, by = 48))] # ages = 18, 20, 24, 28, 32, 36, 40

```

  5. The 10 most frequent sequences

```{r}
# 5 - 10 most frequent sequences

seqtab(df.seq)

```

  6. Plot of all the sequences

```{r}
# 6 - Plot of sequences

seqIplot(df.seq)
```

# 2. Calculate life course dissimilarities in the data

To compute the life course dissimilarities based on the data we are using, we applied a data-driven approach to compute the indel and similarity matrix, based on the transition matrix to assign the transition costs.

```{r dissimilarity,eval=FALSE}
# we will use a not arbitrary indel value based on the transition matrix
trcost <- seqcost(df.seq, method="TRATE")

# calculating distance matrix

dist.om <- seqdist(
  df.seq,
  method="OM",
  indel=trcost$indel,
  sm=trcost$sm
)
```

# 3. Use a clustering methods to derive typical trajectories and describe groups

First, let's analyze the clustering distance distribution based on different number of clusters.

```{r,eval=FALSE}

# aglomerative nesting. it is a way to create a cluster based on different methods

clusterward <-  agnes(
  dist.om, # dissimilarity matrix
  diss = T, # start with dissimilarity matrix
  method = "ward" # method to calculate it
)

```

```{r}
# dendrogram analysis

wardClust <- hclust(as.dist(dist.om), method = "ward.D2")


# statistical indicators

avgClustQual <- as.clustrange(
  clusterward,
  diss=dist.om,
  weights = df$weight40,
  ncluster=20
)

```

```{r plot,include=FALSE}
## plots

# dendro

dendro_plot <- ggdendrogram(
  data = wardClust, rotate = TRUE
) +
  theme(
    axis.text.y = element_blank()
  )

# statistics

indicators_plot <- avgClustQual$stats |>
  as_tibble() |>
  mutate(
    cluster = as.factor(str_remove(rownames(avgClustQual$stats),"cluster")),
    cluster = fct_reorder(cluster,2:20)
  ) |>
  select(any_of(c("R2sq","ASW","HC","HG","PBC","cluster"))) |>
  pivot_longer(
    1:5,
    values_to = "ind_value",
    names_to = "ind_name"
  ) |>
  ggplot() +
  aes(x = cluster, y = ind_value, group = ind_name, color = ind_name) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set3") +
  labs(
    y = "Indicator"
  ) +
  theme_light() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

# first derivative

indicators2_plot <- avgClustQual$stats |>
  as_tibble() |>
  mutate(
    cluster = as.factor(str_remove(rownames(avgClustQual$stats),"cluster")),
    cluster = fct_reorder(cluster,2:20)
  ) |>
  select(any_of(c("R2sq","ASW","HC","HG","PBC","cluster"))) |>
  pivot_longer(
    1:5,
    values_to = "ind_value",
    names_to = "ind_name"
  ) |>
  arrange(ind_name, cluster) |>
  mutate(
    ind_value_lag = lead(ind_value),
    diff_value = ind_value_lag - ind_value,
    diff_value = lag(diff_value),
    .by = c(ind_name)
  ) |>
  filter(!is.na(diff_value)) |>
  ggplot() +
  aes(x = cluster, y = diff_value, group = ind_name, color = ind_name) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set3") +
  labs(
    y = "First derivative of Indicators"
  ) +
  theme_light() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

plot_dend_stat <- dendro_plot + indicators_plot + indicators2_plot

plot_dend_stat
```

Based on the statistic indicators PBC and HG, we the best number of clusters to be chosen is 5.

```{r choosing-number-of-clusters}


clusters5 = cutree(clusterward, k = 5) # choosing the number of clusters

df$clusters <- clusters5

```

```{r,include=FALSE}


# typical trajectories
dc = disscenter(dist.om, group= clusters5, medoids.index="first")
medoid_seq <- print(df.seq[dc,], format = "SPS")
medoid_seq[[1]]

# levels for clusters

# giving a label to these clusters

levels(clusters5)=c(
  medoid_seq[[1]],
  medoid_seq[[2]],
  medoid_seq[[3]],
  medoid_seq[[4]],
  medoid_seq[[5]]
)

```

To describe the groups, let's plot the distribution of trajectories, based on each cluster.

```{r}

# all the sequences for each cluster
seqIplot(
  df.seq,
  group=clusters5,
  border=NA,
  space=0,
  main = levels(clusters5)
)

```


#4. 

For describing group differences in the distribution of life course by cluster, we can use different measures. Here we will use two of them: duration on each state and distribution of states over life course.

  - Duration on each state by cluster

```{r}

# duration of each cluster

seqmtplot(
  df.seq,
  group = clusters5,
  border=NA,
  space=0,
  main = levels(clusters5)
)

```

  - Distribution of states over life course by cluster

```{r}
# distribution of each cluster over observational window
seqdplot(
  df.seq,
  group=clusters5,
  border=NA,
  space=0,
  main = levels(clusters5)
)
```

