---
title: "Indirect estimation Assignment"
author: "Emma & Thiago"
date: "2025-07-07"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#set up ###########################
rm(list = ls())
options(scipen = 9999999)

#Packages ----------------------------
packages <- c(
  "rprojroot", "devtools", "rdhs", "demogsurv",
  "DHS.rates", "tidyverse", "ipumsr", "ggplot2", 
  "readxl", "haven")
purrr::walk(packages, library, character.only = TRUE)

#set wd -------------------------------
project_root <- here::here()
setwd(project_root)

```
### Estimations of adult mortality

*We decided to extract the probability 45q15 based on sibling survival data* and compare it with WPP estimates, based on the last release (UN/DESA, WPP 2024).

```{r adult-mortality,echo=FALSE,warning=FALSE}

# parameter for loop
country_2d <- "GU" # country
type_data <- "SSH"
dhs_years <- c(1995,2015)

for(i in seq_along(dhs_years)){
  # Import data
  load(glue::glue(file.path(project_root,"data/GU/{type_data}{country_2d}{dhs_years[i]}DHS.rda")))

  # rename it
  irfile <- SSH

  # data handling
  irfile$v021 = labelled::unlabelled(irfile$v021)
  irfile$v023 = labelled::unlabelled(irfile$v023)

  # reshaping dataset
  sib <- irfile[grep("caseid|^v0|^v1|^b|^mm", names(irfile))]
  sib = reshape_sib_data(sib, widevars = c("v005", "v008", "v021", "v013", "v024"))
  sib = sib[sib$mm2 %in% c(1,0) & sib$mm1 %in% c(1,2),]
  sib$death <- sib$mm2 == 0; sib$mm8 <- sib$mm8 + 0.5
  sib$mm1 = as.factor(sib$mm1)

  # estimating it
  nqx <- calc_nqx(sib, by=~mm1, agegr=seq(15, 60, 5), tips=c(0, 4, 8, 12), dob="mm4", dod="mm8", strata = NULL) |>
    select("sex" = mm1, everything()) |>
    mutate(
      sex = factor(sex, levels = c(1,2),labels = c("Female","Male"))
    )

  # period of reference

  avg <- mean(sib$v008 / 12 + 1900, na.rm = TRUE) #we convert the CMC into a date + take the mean date (based on days)

  # adding period of reference to dataset

  nqx <- nqx |>
    as_tibble() |>
    mutate(
      referecence_date = avg
    ) |>
    # transforming TIPS to years
    mutate(
      tips_midpoint = case_when(
        tips == "0-3" ~ 1.5,
        tips == "4-7" ~ 5.5,
        tips == "8-11" ~ 10.5
      ),
      reference_year = round(referecence_date - tips_midpoint,0),
      survey = paste0(type_data,country_2d,dhs_years[i])
    )

  # storing results
  if(!"df" %in% ls()){
    df <- nqx
  } else{
    df <- df |>
      bind_rows(nqx)
  }
  rm(nqx)
}




### Compare data with other datasources
# wpp2024 - we have to manipulate age-specific mortality rates to get 35q15
# WDI - world bank data - they use WPP data

# auxiliar table
country_table <- read.delim(
  file = "https://github.com/thiagocalm/tca_utils/raw/refs/heads/master/from-to%20tables/table_countries_uncodes.txt",
  header = TRUE,
  sep = ","
)

# alpha 2 digit codes
alpha2d <- country_table[country_table$Country == "Guatemala",]$Alpha2code

wpp_45q15 <- WDI::WDI(country = alpha2d, indicator = "SP.DYN.AMRT.MA") |>
  select(everything(),"nqx" = 5) |>
  filter(
    !is.na(nqx),
    year %in% 1980:2015
  ) |>
  mutate(
    sex = "Female"
  ) |>
  bind_rows(
    WDI::WDI(country = alpha2d, indicator = "SP.DYN.AMRT.FE") |>
      select(everything(),"nqx" = 5) |>
      filter(
        !is.na(nqx),
        year %in% 1980:2015
      ) |>
      mutate(
        sex = "Male"
      )
  ) |>
  mutate(survey = "WPP 2024 Reivew") |>
  select(survey,year,sex, nqx)


# joining datasets

df <- df |>
  select(survey,"year" = reference_year,sex, "nqx" = est, ci_l, ci_u) |>
  mutate(sex = as.character(sex)) |>
  mutate(
    nqx = nqx * 1000,
    ci_l = ci_l * 1000,
    ci_u = ci_u * 1000
  ) |>
  bind_rows(wpp_45q15)

# plot

ggplot(df, aes(x = year, y = nqx)) +
  geom_ribbon(aes(ymin = ci_l, ymax = ci_u, group = interaction(survey,sex)), fill = "grey88", alpha = 0.1, color = NA, show.legend = FALSE) +
  geom_line(aes(color = survey, linetype = sex), linewidth = 0.8) +
  geom_point(aes(color = survey), size = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = bquote('Adult mortality: '~{""[45]} * q[15]~' by sex in Guatemala' ),
    x = "Reference Year",
    y = bquote(~{""[45]} * q[15]~' per 1000'),
    color = "Data Source",
    linetype = "",
    fill = "Data Source",
    caption = "Data: DHS and UN/DESA, WPP 2024 Revision."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```

Comparing the estimates produced based on the Sibling Survival Histories (SSH) from DHS with those published by the United Nations (WPP, 2024 Review), we observe different trends depending on the year of the DHS survey and sex. First, while the estimates obtained with 2015 DHS show trends for both sexes similar to the ones observed on WPP estimates, trends obtained with 1995 DHS indicate not just a different level but also a different downward trend for both sexes.

These different trends when comparing the survey estimates with WPP ones indicate that some issues regarding the data collection can be affecting the oldest survey. Among the possible issues, selection bias related to the women who we interviewed in the oldest survey can potentially be contributing to an underestimate of mortality.

Similar to the problem often observed on child mortality estimates, the more distant periods we estimate adult mortality, the less reliable these estimates can be due to recall bias. In the case of adult mortality using SSH data, it can be even more worthwhile if we increase the number of siblings someone had over their life. This issue doesn't seem to affect directly our estimates for Male, while it can be affecting the Female estimates, contributing to the lower adult mortality compared to WPP estimates for the most recent survey.

It is valued to report that once we don't have SSH data available for 1999, we can't compare the differences of overlapped estimates. In future analysis, other kinds of indirect estimates can be included to have cross-validation of DHS estimates.
