---
title: "Population Projection"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r packages}
library(pacman)
p_load(tidyverse, wpp2024)
# external functions
source("codes/UNWPP_data.R")
source("codes/UNWPP_download.R")
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer
```

# Exercise 1

We will choose data from the United Nations, World Population Prospects (Revision of 2024) for Calcasian countries.

```{r}
data("pop1")
```

Let's handle with this data to get data just for Calcasian region

```{r ex1_handling_data}

pop1 <- pop1 |>
  as_tibble() |> 
  filter(name %in% c("Azerbaijan","Georgia","Armenia")) |> 
  pivot_longer(`1949`:`2023`, names_to = "year",values_to = "N") |> 
  filter(year %in% c(2009,2014,2019))

```

**1. Estimating population growth rate using different time interval**

To estimate that, we can use the following equation: $$r = \frac{ln(\frac{N(T)}{N(0)})}{T}$$

Where: T is the number of years between 0 and T.

```{r}

# function for calculating the exponential growth rate

PopProj_growthrate <- function(N0,NT,time){
  r = log(NT/N0)/time
}

# Growth rates

growth_rates <- pop1 |> 
  pivot_wider(names_from = year, values_from = N) |> 
  reframe(
    r_5y = PopProj_growthrate(NT = `2019`, N0 = `2014`,time = 5),
    r_10y = PopProj_growthrate(NT = `2019`, N0 = `2009`,time = 10),
    .by = name
  )

```

**2. Project the population 25-years ahead using the constant exponential model and  the two different growth rates**

To project that for each country, we can use the following equation: $$N(t) = N(0) \cdot e^{rt}$$

For that, we will assume different growth rates for each country.

```{r}

pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    .by = name
  )

```

**3. Make an assumption about one of the demographic components, and compare this projection with the baseline one.**

As we know that the growth rate is composed by mortality, fertility and migration, we can rewrite the growth rate based on crude growth rates: $$r = b - d + m$$

Where $b$ is crude birth rate, $d$ is crude death rate, and $m$ is the crude net migration rate.

It is important to point out that our growth rate are still constant. Thus, to keep using the same model - constant growth rate - we have to assume that this change will happen just once.

Let's assume that the growth rates will decrease in 0.08% due to increases in mortality.

```{r}
pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  mutate(
    r_5y_adj = r_5y - 0.0008,
    r_10y_adj = r_10y - 0.0008
  ) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    NProj_25y_r5y_adj = N * exp(r_5y * 25),
    NProj_25y_r10y_adj = N * exp(r_10y * 25),
    .by = name
  )
```

**4. Assume that your observed data ends 10 years before the last two available data point. Project the population 10-year ahead with the two growth rates estimated in point 1. which of the two projections is more accurate?**

```{r}

result = pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2009) |> 
  summarise(
    NProj_2019_r5y = N * exp(r_5y * 10),
    NProj_2019_r10y = N * exp(r_10y * 10),
    .by = name
  ) |> 
  left_join(
    pop1 |> filter(year == 2019) |> select(name, "Pop_observed_2019" = N),
    by = join_by(name)
  )

print(result)

result |> 
  mutate(
    diff_r5y = NProj_2019_r5y - Pop_observed_2019,
    diff_r10y = NProj_2019_r10y - Pop_observed_2019
  )

```

As expected, the one using the growth rates calculated using the 10 years of difference is more accurate once we have the same result than the estimated population in 2019.

# Exercise 2


```{r}

# Downloading the data

download_wpp24(indicator = "mortality")
download_wpp24(indicator = "fertility")

# Importing data from UNWPP

lt_female <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Female",
  indicator = "mortality"
)

lt_male <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Male",
  indicator = "mortality"
)

lt <- bind_rows(lt_female, lt_male) |> 
  filter(year %in% c(2015,2020)) |> 
  select(Location, Sex, year, AgeGrp, age, Lx, lx, mx)

# fertility

fertility_rates <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Female",
  indicator = "fertility"
)

fertility_rates <- fertility_rates |> 
  filter(year %in% c(2015,2020))

# population

library(wpp2024)
data("popAge5dt")

pop <- popAge5dt |> 
  filter(
    name %in% c("Azerbaijan","Georgia","Armenia"),
    year %in% c(2015,2020)
  )

## Handling the data to have 5 years age group
data("age5categories") # let's use this data as an auxiliar one

# mortality
lt <- lt |> 
  select(-AgeGrp) |>
  left_join(
    age5categories |> select(age1,agecat,"AgeGrp" = age),
    by = c("age" = "age1")
  ) |>
  mutate(
    agecat = case_when(agecat >= 85 ~ 85, TRUE ~ agecat),
    AgeGrp = case_when(agecat >= 85 ~ "85+", TRUE ~ AgeGrp)
  ) |>
  reframe(
    mx = mean(mx, na.rm = TRUE),
    lx = lx,
    Lx = sum(Lx),
    .by = c(Location, Sex, year, agecat, AgeGrp)
  ) |>
  mutate(
    Lx = case_when(agecat == 85 ~ lx/mx, TRUE ~ Lx)
  ) |> 
  arrange(Location, Sex, year, agecat, desc(lx)) |> 
  mutate(
    rank = row_number(),
    .by = c(Location, Sex, year, agecat)
  ) |> 
  filter(rank == 1) |> 
  select(-rank, -mx, -lx) |> 
  pivot_wider(names_from = Sex, values_from = Lx) |> 
  select(everything(), "LxF" = Female, "LxM" = Male)
      
# fertility
fertility_rates <- fertility_rates |> 
  left_join(
    age5categories |> select(age1,agecat,"AgeGrp" = age),
    by = c("age" = "age1")
  ) |> 
  reframe(
    fx = mean(fx),
    .by = c(Location, year, agecat, AgeGrp)
  )

# putting everything

df <- pop |> 
  as_tibble() |>
  select("Location" = name, year, age,popM,popF) |> 
  left_join(
    age5categories |> select(agecat,"AgeGrp" = age),
    by = c("age" = "AgeGrp")
  ) |>
  distinct() |> 
  mutate(
    agecat = case_when(agecat >= 85 ~ 85, TRUE ~ agecat),
    age = case_when(agecat >= 85 ~ "85+", TRUE ~ age)
  ) |> 
  reframe(
    popM = sum(popM),
    popF = sum(popF),
    .by = c(Location, year, age, agecat)
  ) |> 
  left_join(
    lt,
    by = join_by(Location, year, agecat)
  ) |> 
  select(-AgeGrp) |> 
  left_join(
    fertility_rates,
    by = join_by(Location, year, agecat)
  ) |> 
    select(-AgeGrp) |> 
  mutate(across(everything(), ~ replace_na(., 0)))

```

Projecting the population from 2015 to 2020.

```{r}
## function to project several periods
PopProj_Leslei <- function(x,AgeGroup,NFx,sFx,bFx,NMx,sMx,bMx,n){
  if(is.na(bFx[length(x)])){
    bFx[length(x)] <- 0
  }
  if(is.na(bMx[length(x)])){
    bMx[length(x)] <- 0
  }
  if(is.na(sFx[length(x)])){
    sFx <- sFx[-length(sFx)]
  }
  if(is.na(sMx[length(x)])){
    sMx <- sFx[-length(sMx)]
  }
  ## number of age groups
  m <- length(x); m2 <- m*2
  ## female Leslie matrix
  LF <- matrix(0,m,m)
  LF[1,] <- bFx
  diag(LF[-1,]) <- sFx
  LF[m,m] <- sFx[m-1]
  ## male Leslie matrix
  BM <- LM <- matrix(0,m,m)
  BM[1,] <- bMx
  diag(LM[-1,]) <- sMx
  LM[m,m] <- sMx[m-1]
  ## putting them together
  ZEROS <- diag(0,m)
  Lup <- cbind(LF,ZEROS)
  Ldown <- cbind(BM,LM)
  L <- rbind(Lup,Ldown)
  ## create population matrix
  N <- matrix(0,m2,n+1)
  N[,1] <- c(NFx,NMx)
  for (i in 1:n){
    N[,i+1] <- L%*%N[,i]
  }
  out <- cbind(
    data.frame(
      x=rep(x,2),
      AgeGroup=rep(AgeGroup,2),
      sex=rep(c("Females","Males"),each=m)
    ),N
  )
return(out)
}

# setting parameters for population projection
SRB <- 1.05 # defining the sex ratio at birth
fact.SRB.F <- 1/(1+SRB)
fact.SRB.M <- SRB/(1+SRB)
l0 <- 100000
L0 <- df |> 
  filter(year == 2015, agecat == 0) |> 
  select(Location, "LF0" = LxF, "LM0" = LxM)

# creating survived births
df_2015 <- df %>%
  filter(year == 2015) |>
  left_join(L0, by = join_by(Location)) |> 
  mutate(
    # Female survivorship ratio
    SxF=lead(LxF)/LxF,
    SxF=case_when (agecat != 80 ~ SxF,
                   agecat == 80 ~ lead(LxF)/(LxF + lead(LxF))),
    # male survivorship ratio
    SxM=lead(LxM)/LxM,
    SxM=case_when (agecat != 80 ~ SxM,
                   agecat == 80 ~ lead(LxM)/(LxM + lead(LxM))),
    # survived births by sex
    bxF=fact.SRB.F * LF0 / (2*l0) * (fx + SxF*lead(fx)),
    bxM=fact.SRB.M * LM0 / (2*l0) * (fx + SxF*lead(fx)),
    .by = c(Location, year)
  )

## Projecting population from 2015 to 2020

# looping it
PopProj = NULL
countries = df_2015$Location |> unique()
for(i in seq_along(countries)){
  cnt = countries[i] # selecting the country name for this time
  proj <- PopProj_Leslei(
    x=df_2015$agecat[df_2015$Location == cnt],
    AgeGroup=df_2015$age[df_2015$Location == cnt],
    NFx=df_2015$popF[df_2015$Location == cnt],
    sFx=df_2015$SxF[df_2015$Location == cnt],
    bFx=df_2015$bxF[df_2015$Location == cnt],
    NMx=df_2015$popM[df_2015$Location == cnt],
    sMx=df_2015$SxM[df_2015$Location == cnt],
    bMx=df_2015$bxM[df_2015$Location == cnt],
    n=1
  )
  
  if(i == 1){
    PopProj <- cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  } else{
    PopProj <- rbind(
      PopProj,
      cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  )
  }
}

# plot

# projected
PopProj |> 
  as_tibble() |> 
  rename(`2015` = `1`, `2020` = `2`) |> 
  pivot_longer(`2015`:`2020`, names_to = "year",values_to = "pop") |> 
  ggplot() +
  aes(x = x, y = pop, color = as.factor(year)) +
  geom_point(size = 3, alpha = .5) +
  geom_line(linewidth = 1.3) +
  lemon::facet_rep_grid(sex~country,repeat.tick.labels = TRUE) +
  theme_light()

# projected x observed

df_2020 <- df |> 
  filter(year == 2020) |> 
  select("country" = Location, year, "x" = agecat, "Males" = popM, "Females" = popF) |> 
  pivot_longer(Males:Females, names_to = "sex", values_to = "EstimatedPop")
  

PopProj |> 
  as_tibble() |> 
  rename(`2015` = `1`, `2020` = `2`) |> 
  pivot_longer(`2015`:`2020`, names_to = "year",values_to = "ProjectedPop") |>
  filter(year == 2020) |> 
  left_join(
    df_2020,
    by = join_by(country, x, sex)
  ) |> 
  pivot_longer(c(ProjectedPop,EstimatedPop), names_to = "type", values_to = "N") |> 
  ggplot() +
  aes(x = x, y = N, color = as.factor(type)) +
  geom_point(size = 3, alpha = .5) +
  geom_line(linewidth = 1.3) +
  lemon::facet_rep_grid(sex~country,repeat.tick.labels = TRUE) +
  theme_light()


```

