---
title: "Population Projection"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r packages}
library(pacman)
p_load(tidyverse, wpp2024)
# external functions
source("codes/UNWPP_data.R")
source("codes/UNWPP_download.R")
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer
```

# Exercise 1

We will choose data from the United Nations, World Population Prospects (Revision of 2024) for Calcasian countries.

```{r}
data("pop1")
```

Let's handle with this data to get data just for Calcasian region

```{r ex1_handling_data}

pop1 <- pop1 |>
  as_tibble() |> 
  filter(name %in% c("Azerbaijan","Georgia","Armenia")) |> 
  pivot_longer(`1949`:`2023`, names_to = "year",values_to = "N") |> 
  filter(year %in% c(2009,2014,2019))

```

**1. Estimating population growth rate using different time interval**

To estimate that, we can use the following equation: $$r = \frac{ln(\frac{N(T)}{N(0)})}{T}$$

Where: T is the number of years between 0 and T.

```{r}

# function for calculating the exponential growth rate

PopProj_growthrate <- function(N0,NT,time){
  r = log(NT/N0)/time
}

# Growth rates

growth_rates <- pop1 |> 
  pivot_wider(names_from = year, values_from = N) |> 
  reframe(
    r_5y = PopProj_growthrate(NT = `2019`, N0 = `2014`,time = 5),
    r_10y = PopProj_growthrate(NT = `2019`, N0 = `2009`,time = 10),
    .by = name
  )

```

**2. Project the population 25-years ahead using the constant exponential model and  the two different growth rates**

To project that for each country, we can use the following equation: $$N(t) = N(0) \cdot e^{rt}$$

For that, we will assume different growth rates for each country.

```{r}

pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    .by = name
  )

```

**3. Make an assumption about one of the demographic components, and compare this projection with the baseline one.**

As we know that the growth rate is composed by mortality, fertility and migration, we can rewrite the growth rate based on crude growth rates: $$r = b - d + m$$

Where $b$ is crude birth rate, $d$ is crude death rate, and $m$ is the crude net migration rate.

It is important to point out that our growth rate are still constant. Thus, to keep using the same model - constant growth rate - we have to assume that this change will happen just once.

Let's assume that the growth rates will decrease in 0.08% due to increases in mortality.

```{r}
pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  mutate(
    r_5y_adj = r_5y - 0.0008,
    r_10y_adj = r_10y - 0.0008
  ) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    NProj_25y_r5y_adj = N * exp(r_5y * 25),
    NProj_25y_r10y_adj = N * exp(r_10y * 25),
    .by = name
  )
```

**4. Assume that your observed data ends 10 years before the last two available data point. Project the population 10-year ahead with the two growth rates estimated in point 1. which of the two projections is more accurate?**

```{r}

result = pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2009) |> 
  summarise(
    NProj_2019_r5y = N * exp(r_5y * 10),
    NProj_2019_r10y = N * exp(r_10y * 10),
    .by = name
  ) |> 
  left_join(
    pop1 |> filter(year == 2019) |> select(name, "Pop_observed_2019" = N),
    by = join_by(name)
  )

print(result)

result |> 
  mutate(
    diff_r5y = NProj_2019_r5y - Pop_observed_2019,
    diff_r10y = NProj_2019_r10y - Pop_observed_2019
  )

```

As expected, the one using the growth rates calculated using the 10 years of difference is more accurate once we have the same result than the estimated population in 2019.

# Exercise 2


```{r}

# Downloading the data

download_wpp24(indicator = "mortality")
download_wpp24(indicator = "fertility")

# Importing data from UNWPP

lt_female <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2014,
  end_year = 2019,
  sex = "Female",
  indicator = "mortality"
)

lt_male <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2014,
  end_year = 2019,
  sex = "Male",
  indicator = "mortality"
)

lt <- bind_rows(lt_female, lt_male)
rm(lt_female, lt_male)

# fertility

fertility_rates <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2014,
  end_year = 2019,
  sex = "Female",
  indicator = "fertility"
)

# population
pop_male <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2014,
  end_year = 2019,
  sex = "Male",
  indicator = "population"
)

pop_female <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2014,
  end_year = 2019,
  sex = "Male",
  indicator = "population"
)

```

