---
title: "Population Projection"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r packages}
library(pacman)
p_load(tidyverse, wpp2024)
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer
```

# Exercise 1

We will choose data from the United Nations, World Population Prospects (Revision of 2024) for Calcasian countries.

```{r}
data("pop1")
```

Let's handle with this data to get data just for Calcasian region

```{r ex1_handling_data}

pop1 <- pop1 |>
  as_tibble() |> 
  filter(name %in% c("Azerbaijan","Georgia","Armenia")) |> 
  pivot_longer(`1949`:`2023`, names_to = "year",values_to = "N") |> 
  filter(year %in% c(2009,2014,2019))

```

**1. Estimating population growth rate using different time interval**

To estimate that, we can use the following equation: $$r = \frac{ln(\frac{N(T)}{N(0)})}{T}$$

Where: T is the number of years between 0 and T.

```{r}

# function for calculating the exponential growth rate

PopProj_growthrate <- function(N0,NT,time){
  r = log(NT/N0)/time
}

# Growth rates

growth_rates <- pop1 |> 
  pivot_wider(names_from = year, values_from = N) |> 
  
  reframe(
    r_5y = PopProj_growthrate(NT = `2019`, N0 = `2014`,time = 5)*100,
    r_10y = PopProj_growthrate(NT = `2019`, N0 = `2009`,time = 10)*100,
    .by = name
  )



```

**2. Project the population 25-years ahead using the constant exponential model and  the two differentgrowth rates**
