---
title: "Population Projection"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r packages}
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer

# packages
library(pacman)
p_load(tidyverse, wpp2024)

# external functions
source("codes/UNWPP_data.R")
source("codes/UNWPP_download.R")
```

# Exercise 1

We will choose data from the United Nations, World Population Prospects (Revision of 2024) for Calcasian countries.

```{r}
data("pop1")
```

Let's handle with this data to get data just for Calcasian region

```{r ex1_handling_data}

pop1 <- pop1 |>
  as_tibble() |> 
  filter(name %in% c("Azerbaijan","Georgia","Armenia")) |> 
  pivot_longer(`1949`:`2023`, names_to = "year",values_to = "N") |> 
  filter(year %in% c(2009,2014,2019))

```

**1. Estimating population growth rate using different time interval**

To estimate that, we can use the following equation: $$r = \frac{ln(\frac{N(T)}{N(0)})}{T}$$

Where: T is the number of years between 0 and T.

```{r}

# function for calculating the exponential growth rate

PopProj_growthrate <- function(N0,NT,time){
  r = log(NT/N0)/time
}

# Growth rates

growth_rates <- pop1 |> 
  pivot_wider(names_from = year, values_from = N) |> 
  reframe(
    r_5y = PopProj_growthrate(NT = `2019`, N0 = `2014`,time = 5),
    r_10y = PopProj_growthrate(NT = `2019`, N0 = `2009`,time = 10),
    .by = name
  )

growth_rates

```

**2. Project the population 25-years ahead using the constant exponential model and  the two different growth rates**

To project that for each country, we can use the following equation: $$N(t) = N(0) \cdot e^{rt}$$

For that, we will assume different growth rates for each country.

```{r}

pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    .by = name
  )

```

**3. Make an assumption about one of the demographic components, and compare this projection with the baseline one.**

As we know that the growth rate is composed by mortality, fertility and migration, we can rewrite the growth rate based on crude growth rates: $$r = b - d + m$$

Where $b$ is crude birth rate, $d$ is crude death rate, and $m$ is the crude net migration rate.

It is important to point out that our growth rate are still constant. Thus, to keep using the same model - constant growth rate - we have to assume that this change will happen just once.

Let's assume that the growth rates will decrease in 0.08% due to increases in mortality.

```{r}
pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2019) |> 
  mutate(
    r_5y_adj = r_5y - 0.0008,
    r_10y_adj = r_10y - 0.0008
  ) |> 
  summarise(
    NProj_25y_r5y = N * exp(r_5y * 25),
    NProj_25y_r10y = N * exp(r_10y * 25),
    NProj_25y_r5y_adj = N * exp(r_5y * 25),
    NProj_25y_r10y_adj = N * exp(r_10y * 25),
    .by = name
  )
```

**4. Assume that your observed data ends 10 years before the last two available data point. Project the population 10-year ahead with the two growth rates estimated in point 1. which of the two projections is more accurate?**

```{r}

result = pop1 |> 
  left_join(
    growth_rates,
    by = join_by(name)
  ) |> 
  filter(year == 2009) |> 
  summarise(
    NProj_2019_r5y = N * exp(r_5y * 10),
    NProj_2019_r10y = N * exp(r_10y * 10),
    .by = name
  ) |> 
  left_join(
    pop1 |> filter(year == 2019) |> select(name, "Pop_observed_2019" = N),
    by = join_by(name)
  )

result |> 
  mutate(
    diff_r5y = NProj_2019_r5y - Pop_observed_2019,
    diff_r10y = NProj_2019_r10y - Pop_observed_2019
  )

```

As expected, the one using the growth rates calculated using the 10 years of difference is more accurate once we have the same result than the estimated population in 2019.

# Exercise 2

**Take the population for any country (for example, use data from the HMD), as well as data for person-years and fertility rates (for example, take data from HMD and HFD). Specifically, take these data 5 years before the last available date (e.g. if the last available year is 2020, take the data in 2015). Divide the population into 5-year age groups and project the population 5-year ahead for each sex separately. Plot and compare your results with the observed population in the last available data (e.g. in 2020). Are your projections close to the observed population? At what ages do you see larger differences? Briefly provide a motivation for this discrepancy**

We will use the World Population Prospects (Revision 2024) data for Caucasus countries (Armenia, Azerbaijan and Georgia). For that, we will choose as last period the year of 2020 and we will project the population from 2015 to 2020.

Once it is not specified the method of projection in the exercise prompt, we chose Leslei matrix approach due to the easier way to be implemented comparing to cohort-component method (CCM).

```{r}

# Downloading data from wpp website to have life table and age-specific fertility rates

download_wpp24(indicator = "mortality")
download_wpp24(indicator = "fertility")

# Life tables from wpp website

lt_female <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Female",
  indicator = "mortality"
)

lt_male <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Male",
  indicator = "mortality"
)

lt <- bind_rows(lt_female, lt_male) |> 
  filter(year %in% c(2015,2020)) |> 
  select(Location, Sex, year, AgeGrp, age, Lx, lx, mx)

# age-specific fertility rates from WPP website

fertility_rates <- UNWPP_data(
  country = c("Azerbaijan","Georgia","Armenia"),
  start_year = 2015,
  end_year = 2020,
  sex = "Female",
  indicator = "fertility"
)

fertility_rates <- fertility_rates |> 
  filter(year %in% c(2015,2020))

# population

data("popAge5dt") # importing data from WPP (wpp2024 package)

pop <- popAge5dt |> 
  filter(
    name %in% c("Azerbaijan","Georgia","Armenia"),
    year %in% c(2015,2020)
  )

## Handling the data to have 5 years age group
data("age5categories") # let's use this dataset as an auxiliary one

# mortality
lt <- lt |> 
  select(-AgeGrp) |>
  # join auxiliary dataset with description of age groups
  left_join(
    age5categories |> select(age1,agecat,"AgeGrp" = age),
    by = c("age" = "age1")
  ) |>
  # changing last age group to become 85+
  mutate(
    agecat = case_when(agecat >= 85 ~ 85, TRUE ~ agecat),
    AgeGrp = case_when(agecat >= 85 ~ "85+", TRUE ~ AgeGrp)
  ) |>
  # summing up values for 5-years age group
  reframe(
    mx = mean(mx, na.rm = TRUE),
    lx = lx,
    Lx = sum(Lx),
    .by = c(Location, Sex, year, agecat, AgeGrp)
  ) |>
  # adjusting for last age group data
  mutate(
    Lx = case_when(agecat == 85 ~ lx/mx, TRUE ~ Lx)
  ) |> 
  # adjusts to have a dataframe with just one value for each age group
  arrange(Location, Sex, year, agecat, desc(lx)) |> 
  mutate(
    rank = row_number(),
    .by = c(Location, Sex, year, agecat)
  ) |> 
  filter(rank == 1) |> 
  # excluding undesired variables
  select(-rank, -mx, -lx) |> 
  pivot_wider(names_from = Sex, values_from = Lx) |> 
  select(everything(), "LxF" = Female, "LxM" = Male)
      
# fertility
fertility_rates <- fertility_rates |> 
  # join auxiliary dataset with description of age groups
  left_join(
    age5categories |> select(age1,agecat,"AgeGrp" = age),
    by = c("age" = "age1")
  ) |> 
  # taking the average of asfr for each group
  reframe(
    fx = mean(fx),
    .by = c(Location, year, agecat, AgeGrp)
  )

# putting everything

df <- pop |> 
  as_tibble() |>
  select("Location" = name, year, age,popM,popF) |> 
  # adding auxiliary dataset with age groups
  left_join(
    age5categories |> select(agecat,"AgeGrp" = age),
    by = c("age" = "AgeGrp")
  ) |>
  distinct() |> 
  mutate(
    agecat = case_when(agecat >= 85 ~ 85, TRUE ~ agecat),
    age = case_when(agecat >= 85 ~ "85+", TRUE ~ age)
  ) |> 
  # aggregating population in 5-years age groups
  reframe(
    popM = sum(popM),
    popF = sum(popF),
    .by = c(Location, year, age, agecat)
  ) |> 
  # join LT dataset
  left_join(
    lt,
    by = join_by(Location, year, agecat)
  ) |> 
  select(-AgeGrp) |> 
  # join ASFR dataset
  left_join(
    fertility_rates,
    by = join_by(Location, year, agecat)
  ) |> 
    select(-AgeGrp) |> 
  mutate(across(everything(), ~ replace_na(., 0)))

```

Projecting the population from 2015 to 2020 for each of these countries.

```{r}
## function to project several periods
PopProj_Leslei <- function(x,AgeGroup,NFx,sFx,bFx,NMx,sMx,bMx,n){
  if(is.na(bFx[length(x)])){
    bFx[length(x)] <- 0
  }
  if(is.na(bMx[length(x)])){
    bMx[length(x)] <- 0
  }
  if(is.na(sFx[length(x)])){
    sFx <- sFx[-length(sFx)]
  }
  if(is.na(sMx[length(x)])){
    sMx <- sFx[-length(sMx)]
  }
  ## number of age groups
  m <- length(x); m2 <- m*2
  ## female Leslie matrix
  LF <- matrix(0,m,m)
  LF[1,] <- bFx
  diag(LF[-1,]) <- sFx
  LF[m,m] <- sFx[m-1]
  ## male Leslie matrix
  BM <- LM <- matrix(0,m,m)
  BM[1,] <- bMx
  diag(LM[-1,]) <- sMx
  LM[m,m] <- sMx[m-1]
  ## putting them together
  ZEROS <- diag(0,m)
  Lup <- cbind(LF,ZEROS)
  Ldown <- cbind(BM,LM)
  L <- rbind(Lup,Ldown)
  ## create population matrix
  N <- matrix(0,m2,n+1)
  N[,1] <- c(NFx,NMx)
  for (i in 1:n){
    N[,i+1] <- L%*%N[,i]
  }
  out <- cbind(
    data.frame(
      x=rep(x,2),
      AgeGroup=rep(AgeGroup,2),
      sex=rep(c("Females","Males"),each=m)
    ),N
  )
return(out)
}

# setting parameters for population projection
SRB <- 1.05 # defining the sex ratio at birth
fact.SRB.F <- 1/(1+SRB)
fact.SRB.M <- SRB/(1+SRB)
l0 <- 100000
L0 <- df |> 
  filter(year == 2015, agecat == 0) |> 
  select(Location, "LF0" = LxF, "LM0" = LxM)

# creating survived births
df_2015 <- df %>%
  filter(year == 2015) |>
  left_join(L0, by = join_by(Location)) |> 
  mutate(
    # Female survivorship ratio
    SxF=lead(LxF)/LxF,
    SxF=case_when (agecat != 80 ~ SxF,
                   agecat == 80 ~ lead(LxF)/(LxF + lead(LxF))),
    # male survivorship ratio
    SxM=lead(LxM)/LxM,
    SxM=case_when (agecat != 80 ~ SxM,
                   agecat == 80 ~ lead(LxM)/(LxM + lead(LxM))),
    # survived births by sex
    bxF=fact.SRB.F * LF0 / (2*l0) * (fx + SxF*lead(fx)),
    bxM=fact.SRB.M * LM0 / (2*l0) * (fx + SxF*lead(fx)),
    .by = c(Location, year)
  )

## Projecting population from 2015 to 2020

# looping it
PopProj = NULL
countries = df_2015$Location |> unique()
for(i in seq_along(countries)){
  cnt = countries[i] # selecting the country name for this time
  proj <- PopProj_Leslei(
    x=df_2015$agecat[df_2015$Location == cnt],
    AgeGroup=df_2015$age[df_2015$Location == cnt],
    NFx=df_2015$popF[df_2015$Location == cnt],
    sFx=df_2015$SxF[df_2015$Location == cnt],
    bFx=df_2015$bxF[df_2015$Location == cnt],
    NMx=df_2015$popM[df_2015$Location == cnt],
    sMx=df_2015$SxM[df_2015$Location == cnt],
    bMx=df_2015$bxM[df_2015$Location == cnt],
    n=1
  )
  
  if(i == 1){
    PopProj <- cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  } else{
    PopProj <- rbind(
      PopProj,
      cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  )
  }
}

```

Now, let's evaluate the results comparing the baseline population (2015) and the projected one for 2020.

```{r}
# plot

# projected
PopProj |> 
  as_tibble() |> 
  rename(`2015` = `1`, `2020` = `2`) |> 
  pivot_longer(`2015`:`2020`, names_to = "year",values_to = "pop") |> 
  ggplot() +
  aes(x = x, y = pop, color = as.factor(year)) +
  geom_point(size = 3, alpha = .5) +
  geom_line(linewidth = 1.3) +
  labs(
    color = "",
    y = "Population (x 1,000)",
    x = "5-years age group",
    caption = "Source: United Nations, DESA. World Population Prospects, Revision 2024.",
    title = "Estimated (2015) and forwards projected population (2020) for Caucasian countries"
  ) +
  scale_color_brewer(palette = "Set2") +
  lemon::facet_rep_grid(sex~country,repeat.tick.labels = TRUE, scales = "free_y") +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", colour = "grey44"),
    axis.title = element_text(face = "bold", colour = "grey44"),
    strip.text = element_text(face = "bold", colour = "grey44"),
    axis.text = element_text(colour = "grey44"),
    plot.caption = element_text(colour = "grey44", hjust = 1),
    legend.position = "bottom"
  )

```

It seems that the projected population follows the observed trends in the baseline population. When we analyse the age structure for each country and sex, we conclude that this echoes observed in the baseline populations are still being observed in the projected ones.

Now we can compare these results with the estimated population in WPP 2024 revision. First, let's assess the comparison in absolute terms.

```{r}
# projected x observed

df_2020 <- df |> 
  filter(year == 2020) |> 
  select("country" = Location, year, "x" = agecat, "Males" = popM, "Females" = popF) |> 
  pivot_longer(Males:Females, names_to = "sex", values_to = "EstimatedPop")

PopProj |>
  as_tibble() |>
  rename(`2015` = `1`, `2020` = `2`) |>
  pivot_longer(`2015`:`2020`, names_to = "year",values_to = "ProjectedPop") |>
  filter(year == 2020) |>
  left_join(
    df_2020,
    by = join_by(country, x, sex)
  ) |>
  pivot_longer(c(ProjectedPop,EstimatedPop), names_to = "type", values_to = "N") |>
  # filter(country != "Azerbaijan") |>
  ggplot() +
  aes(x = x, y = N, color = as.factor(type)) +
  geom_point(size = 3, alpha = .5) +
  geom_line(linewidth = 1.3) +
  labs(
    color = "",
    y = "Population (x 1,000)",
    x = "5-years age group",
    caption = "Source: United Nations, DESA. World Population Prospects, Revision 2024.",
    title = "Estimated (2020) and forwards projected population (2020) for Caucasian countries"
  ) +
  scale_color_brewer(palette = "Set2") +
  lemon::facet_rep_grid(sex~country,repeat.tick.labels = TRUE, scales = "free_y") +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", colour = "grey44"),
    axis.title = element_text(face = "bold", colour = "grey44"),
    strip.text = element_text(face = "bold", colour = "grey44"),
    axis.text = element_text(colour = "grey44"),
    plot.caption = element_text(colour = "grey44", hjust = 1),
    legend.position = "bottom"
  )

```

Overall, we observe similar age composition and size in respect to the projected population and estimated one. In Azerbaijan, absolute differences can be standed out due to the fact they have the biggest population size.

Another way to assess the consistency of the population projection without differences due to population size is based on the ratio between projected population in 2020 and the estimated population for the same calendar year (2020) from WPP estimates. The figure below show this relationship between both values for each country and sex.

```{r}
PopProj |>
  as_tibble() |>
  rename(`2015` = `1`, `2020` = `2`) |>
  pivot_longer(`2015`:`2020`, names_to = "year",values_to = "ProjectedPop") |>
  filter(year == 2020) |>
  left_join(
    df_2020,
    by = join_by(country, x, sex)
  ) |> 
  mutate(ratio = ProjectedPop/EstimatedPop) |> 
  ggplot() +
  aes(x = x, y = ratio, color = as.factor(country)) +
  geom_point(size = 3, alpha = .5) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 1, color = "black",linetype = "dashed", linewidth = 1.3) +
  labs(
    color = "",
    y = "[Projected Population, 2020] / [WPP Estimated Population, 2020]",
    x = "5-years age group",
    caption = "Source: United Nations, DESA. World Population Prospects, Revision 2024.",
    title = "Ratio between projected and estimated population (2020) for Caucasian countries"
  ) +
  scale_color_brewer(palette = "Set2") +
  lemon::facet_rep_grid(.~sex,repeat.tick.labels = TRUE, scales = "free_y") +
  theme_light() +
  theme(
    plot.title = element_text(face = "bold", colour = "grey44"),
    axis.title = element_text(face = "bold", colour = "grey44"),
    strip.text = element_text(face = "bold", colour = "grey44"),
    axis.text = element_text(colour = "grey44"),
    plot.caption = element_text(colour = "grey44", hjust = 1),
    legend.position = "bottom"
  )
```

Two main aspects can be standed out based on the ratio between these populations. Firstly, we see that our projection is overestimating the number of children between 0-4 years old. Once we are keeping constant the fertility rates observed in 2015, we are not taking into account the recent trends observed regarding this component for these countries. In addition, a share of this difference observed can also be affected by Covid-19 effects on births in 2020, contributing for the overestimation of our projection.

Secondly, We can see a higher difference between projected population and the estimated one in old ages. On the one hand, this difference is due to the changes in mortality observed between 2015-2020. Once mortality has kept constant in this projection interval, it would be expected that the projected population would be higher in 2020 than the estimated one by UN. However, for the female population we see a different trend, with lower population for estimated one comparing to the one projected. Among other factors, it can be due to the fact we have a small amount of people in old ages in these countries, what can generate these variations.


# Exercise 3

Use Sanan's code here!

# Exercise 5

**Take again the population that you used for Exercise #2. Project the population for 5 years ahead, but this time use matrix formulas. Compare your results with the projections that you obtained in Exercise #2. Are they the same? Then, project your chosen population by sex for n = 20 periods ahead and show your results.**

Once we used the Leslei approach for the exercise 2, in this exercise we will continue the population projection for 20 intervals (100 years ahead).

To project the population from 2015 to 2115, we will use the same R function created in the exercise 2 (we thanks to Marilia, who structured this function). The R implementation is described below.

```{r}
## Projecting population from 2015 to 2115

# looping it
PopProj20 = NULL
countries = df_2015$Location |> unique()
for(i in seq_along(countries)){
  cnt = countries[i] # selecting the country name for this time
  proj <- PopProj_Leslei(
    x=df_2015$agecat[df_2015$Location == cnt],
    AgeGroup=df_2015$age[df_2015$Location == cnt],
    NFx=df_2015$popF[df_2015$Location == cnt],
    sFx=df_2015$SxF[df_2015$Location == cnt],
    bFx=df_2015$bxF[df_2015$Location == cnt],
    NMx=df_2015$popM[df_2015$Location == cnt],
    sMx=df_2015$SxM[df_2015$Location == cnt],
    bMx=df_2015$bxM[df_2015$Location == cnt],
    n=20
  )
  
  if(i == 1){
    PopProj20 <- cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  } else{
    PopProj20 <- rbind(
      PopProj20,
      cbind(
      country = rep(cnt, nrow(proj)),
      proj
    )
  )
  }
}

### Vizualization


```

