---
title: "EDSD 2024–2025: Population Projection's Assignment"
author:
  - "Sanan Abdulayev"
  - "Gilbert Habaasa"
  - "Thiago C. Almeida"
date:
format:
  pdf:
      }
      \pagestyle{fancy}
      \fancyhead[R]{\includegraphics[width=2cm]{edsd_logo.png}}
      \fancyfoot[C]{}
    keep-tex: true
    header-includes: |
      \usepackage{fancyhdr}
      \usepackage{graphicx}
      \usepackage{fvextra}
      % redefine Highlighting para wrap de código
      \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
        commandchars=\\\{\},
        breaklines=true,
        breakanywhere=true,
        fontsize=\small
      \fancyhead[L]{Instructor: Marília Nepomuceno}
editor: visual
editor_options:
  chunk_output_type: console
---

```{r packages}
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer

# packages
library(pacman)
p_load(tidyverse, wpp2024, here, knitr, kableExtra)

```


# Introduction

For this assignment, we will use data from Sweden, with mortality rates from 1771 to 2022 by age and sex.

```{r}
load("data/MortSWE.Rdata")
```

A general trend of these mortality rates from 1950 onwards can be seen in the graphic below.

```{r}
MORT.SWE |> 
  filter(Year >= 1950, Age <= 100) |> 
  ggplot() +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  facet_wrap(.~Sex) +
  theme_minimal(base_size = 10)
```

# Exercise 1

For this exercise, we will work with male's moratlity from 1950 onwards for newborns (age 0).

```{r}
mort_m0 <- MORT.SWE |> 
  filter(
    Year >= 1950,
    Sex == "Male",
    Age == 0
  )
```

Now, let's fit and forecast mortality from 2022 up to 2050 using two different models:

 - a linear model for death rates;
 
 - a generalized linear model for deaths, using exposures as an offset.
 
```{r}

### MODEL 1 - LM for death rates

mod1 <- lm(Rates ~ Year, data = mort_m0)

summary(mod1)

## predicting values and extrapolating it until 2050

# dimensions of the problem
t <- mort_m0$Year
y <- mort_m0$Rates

# fit and focast of the model
tF <- min(t):2050 # creating the rage of years

df_forecast <- tibble(Year = tF) # creating df with rage of years

# forecasting years

yF <- predict(object = mod1, newdata = df_forecast)

# adding it to df

df_forecast <- df_forecast |> 
  bind_cols(yF) |> 
  select(Year, mx_lm = `...2`) |> 
  mutate(
    type = case_when(Year <= max(t) ~ "Predicted", TRUE ~ "Projected")
  )

### MODEL 2 - GLM for deaths with exposure

# dimensions of the problem
t <- mort_m0$Year
y <- mort_m0$Deaths
e <- mort_m0$Exposures
tF <- min(t):2050 # creating the rage of years

# fit the model
mod1.glm <- glm(y ~ Year, data = mort_m0, offset = log(e), family = poisson())

summary(mod1.glm)

# forecasting years

df_forecast <- df_forecast |> 
  mutate(
    mx_glm = exp(coef(mod1.glm)[1] + coef(mod1.glm)[2] * tF)
  ) |> 
  rename(
    "LM" = mx_lm,
    "GLM" = mx_glm
  ) |> 
  pivot_longer(
    cols = c(LM, GLM),
    values_to = "mx",
    names_to = "model"
  )

# plotting it...

ggplot() +
  geom_line(
    data = df_forecast,
    aes(x = Year, y = mx, linetype = type, color = model),
    linewidth = 1.3
  ) +
  geom_point(
    data = mort_m0,
    aes(x = Year, y = Rates),
    color = "black",
    size = 2
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  theme_minimal()


```

[write the interpretation here...]

# Exercise 2

For this exercise, we will use not only the age 0, but all the range age between 0 and 100.

```{r}
mort <- MORT.SWE |> 
  filter(
    Year >= 1950,
    Sex == "Male",
    between(Age, 0, 100)
  )
```

Let's fit a GLM model for deaths, using an offset, for all age groups available.

```{r}

### MODEL - GLM for deaths with exposure

# dimensions of the problem
t <- unique(mort$Year)
tF <- min(t):2050 # creating the rage of years
age <- unique(mort$Age)
n <- length(age)
m <- length(tF)

# creating object for forecasting
mF <- matrix(data = NA, nrow = n, ncol = m)

# loop for predicted and forecasting mortality rates
i = 1
for(i in seq_along(age)){
  # setting remain dimensions of the problem
  y <- mort$Deaths[mort$Age == age[i]]
  e <- mort$Exposures[mort$Age == age[i]]
  # fit the model
  mod.glm <- glm(y ~ t, offset = log(e), family = poisson())
  #predict model
  mF[i,] <- exp(coef(mod.glm)[1] + coef(mod.glm)[2] * tF)
}

```

To compare the predicted values and the observed ones, we can see the panel of plots below.

```{r}

# Plot 1 - observed rates

plot1 <- mort |> 
  ggplot() +
  # aes(x = Year, y = logRates, color = Age, group = interaction(Age, Age)) +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  scale_color_viridis_c(option = "B") +
  labs(title = "Observed rates") +
  theme_light(base_size = 12)

# Plot 2 - predicted rates

colnames(mF) <- 1950:2050

plot1 <- mF |>
  as_tibble()
  ggplot() +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  scale_color_viridis_c(option = "B") +
  labs(title = "Observed rates") +
  theme_light(base_size = 12)

```


