---
title: "EDSD 2024–2025: Population Projection's Assignment"
author:
  - "Sanan Abdulayev"
  - "Gilbert Habaasa"
  - "Thiago C. Almeida"
date:
format:
  pdf:
      }
      \pagestyle{fancy}
      \fancyhead[R]{\includegraphics[width=2cm]{edsd_logo.png}}
      \fancyfoot[C]{}
    keep-tex: true
    header-includes: |
      \usepackage{fancyhdr}
      \usepackage{graphicx}
      \usepackage{fvextra}
      % redefine Highlighting para wrap de código
      \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
        commandchars=\\\{\},
        breaklines=true,
        breakanywhere=true,
        fontsize=\small
      \fancyhead[L]{Instructor: Marília Nepomuceno}
editor: visual
editor_options:
  chunk_output_type: console
---

```{r packages}
# general settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer

# packages
library(pacman)
p_load(tidyverse, wpp2024, here, knitr, kableExtra, patchwork, forecast, tseries, modelsummary)

```


# Introduction

For this assignment, we will use data from Sweden, with mortality rates from 1771 to 2022 by age and sex.

```{r}
load("data/MortSWE.Rdata")
```

A general trend of these mortality rates from 1950 onwards can be seen in the graphic below.

```{r}
MORT.SWE |> 
  filter(Year >= 1950, Age <= 100) |> 
  ggplot() +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  facet_wrap(.~Sex) +
  theme_minimal(base_size = 10)
```

# Exercise 1

For this exercise, we will work with male's moratlity from 1950 onwards for newborns (age 0).

```{r}
mort_m0 <- MORT.SWE |> 
  filter(
    Year >= 1950,
    Sex == "Male",
    Age == 0
  )
```

Now, let's fit and forecast mortality from 2022 up to 2050 using two different models:

 - a linear model for death rates;
 
 - a generalized linear model for deaths, using exposures as an offset.
 
```{r}

### MODEL 1 - LM for death rates

mod1 <- lm(Rates ~ Year, data = mort_m0)

summary(mod1)

## predicting values and extrapolating it until 2050

# dimensions of the problem
t <- mort_m0$Year
y <- mort_m0$Rates

# fit and focast of the model
tF <- min(t):2050 # creating the rage of years

df_forecast <- tibble(Year = tF) # creating df with rage of years

# forecasting years

yF <- predict(object = mod1, newdata = df_forecast)

# adding it to df

df_forecast <- df_forecast |> 
  bind_cols(yF) |> 
  select(Year, mx_lm = `...2`) |> 
  mutate(
    type = case_when(Year <= max(t) ~ "Predicted", TRUE ~ "Projected")
  )

### MODEL 2 - GLM for deaths with exposure

# dimensions of the problem
t <- mort_m0$Year
y <- mort_m0$Deaths
e <- mort_m0$Exposures
tF <- min(t):2050 # creating the rage of years

# fit the model
mod1.glm <- glm(y ~ Year, data = mort_m0, offset = log(e), family = poisson())

summary(mod1.glm)

# forecasting years

df_forecast <- df_forecast |> 
  mutate(
    mx_glm = exp(coef(mod1.glm)[1] + coef(mod1.glm)[2] * tF)
  ) |> 
  rename(
    "LM" = mx_lm,
    "GLM" = mx_glm
  ) |> 
  pivot_longer(
    cols = c(LM, GLM),
    values_to = "mx",
    names_to = "model"
  )

# plotting it...

ggplot() +
  geom_line(
    data = df_forecast,
    aes(x = Year, y = mx, linetype = type, color = model),
    linewidth = 1.3
  ) +
  geom_point(
    data = mort_m0,
    aes(x = Year, y = Rates),
    color = "black",
    size = 2
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  theme_minimal()


```

[write the interpretation here...]

# Exercise 2

For this exercise, we will use not only the age 0, but all the range age between 0 and 100.

```{r}
mort <- MORT.SWE |> 
  filter(
    Year >= 1950,
    Sex == "Male",
    between(Age, 0, 100)
  )
```

Let's fit a GLM model for deaths, using an offset, for all age groups available.

```{r}

### MODEL - GLM for deaths with exposure

# dimensions of the problem
t <- unique(mort$Year)
tF <- min(t):2050 # creating the rage of years
age <- unique(mort$Age)
n <- length(age)
m <- length(tF)

# creating object for forecasting
mF <- matrix(data = NA, nrow = n, ncol = m)

# loop for predicted and forecasting mortality rates
i = 1
for(i in seq_along(age)){
  # setting remain dimensions of the problem
  y <- mort$Deaths[mort$Age == age[i]]
  e <- mort$Exposures[mort$Age == age[i]]
  # fit the model
  mod.glm <- glm(y ~ t, offset = log(e), family = poisson())
  #predict model
  mF[i,] <- exp(coef(mod.glm)[1] + coef(mod.glm)[2] * tF)
}

```

To compare the predicted values and the observed ones, we can see the panel of plots below.

```{r}

# Plot 1 - observed rates

plot1 <- mort |> 
  ggplot() +
  # aes(x = Year, y = logRates, color = Age, group = interaction(Age, Age)) +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  scale_color_viridis_c(option = "B") +
  coord_cartesian(ylim = c(-10.5,0.5)) +
  scale_y_continuous(breaks = seq(-10,0,1)) +
  labs(title = "Observed rates") +
  theme_light(base_size = 12)

# Plot 2 - predicted rates

colnames(mF) <- 1950:2050

plot2 <- mF |>
  as_tibble() |> 
  mutate(
    Age = age
  ) |> 
  pivot_longer(
    `1950`:`2050`,
    names_to = "Year",
    values_to = "mx_hat"
  ) |>
  mutate(Year = as.numeric(Year)) |> 
  filter(Year <= 2022) |> 
  mutate(
    logRates = log(mx_hat)
  ) |> 
  ggplot() +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  scale_color_viridis_c(option = "B") +
  coord_cartesian(ylim = c(-10.5,0.5)) +
  scale_y_continuous(breaks = seq(-10,0,1)) +
  labs(title = "Predicted rates") +
  theme_light(base_size = 12)

# Plot 3 - predicted and projected

plot3 <- mF |>
  as_tibble() |> 
  mutate(
    Age = age
  ) |> 
  pivot_longer(
    `1950`:`2050`,
    names_to = "Year",
    values_to = "mx_hat"
  ) |>
  mutate(Year = as.numeric(Year)) |> 
  filter(Year > 2022) |> 
  mutate(
    logRates = log(mx_hat)
  ) |> 
  ggplot() +
  aes(x = Age, y = logRates, color = Year, group = interaction(Year, Year)) +
  geom_line(alpha = .3) +
  scale_color_viridis_c(option = "D") +
  coord_cartesian(ylim = c(-10.5,0.5)) +
  scale_y_continuous(breaks = seq(-10,0,1)) +
  labs(title = "Forecasted rates") +
  theme_light(base_size = 12)

plot1 + plot2 + plot3

```

[write the interpretation here...]

# Exercise 3

We will work with the same data set, but with male mortality rates from 1950 to 2000 for age 20.

```{r}
mort_m20 <- MORT.SWE |> 
  filter(
    Year >= 1950,
    Sex == "Male",
    Age == 20
  )
```

Let's fit and forecast mortality up to 2022 using two different models

  - a random walk with drift model for log-rates
  
  - the best possible ARIMA model for log-rates

The first exercise is to describe the data, trying to interpret which kind of model will fit better in this case. To do so, we can use ACF plot and test for stationarity (KPSS).

```{r}

### MODEL 1 - Random walk with drift

# dimensions of the problem
t <- mort_m20$Year[mort_m20$Year <= 2000]
y <- mort_m20$logRates[mort_m20$Year <= 2000]
tF <- min(t):2022 # creating the rage of years

# plotting it...

par(mfrow = c(1,3))
# description of the data
plot(t,y)
# ACF
Acf(y)
# PACF
Acf(y,type = "partial")
par(mfrow = c(1,1))

```

It seems that we are dealing with a non-stationary series. To understand better if we can make the serie become startionary, let's take the first difference of the rates and plot the same graphs again.

```{r}
# taking the first-order diff...
y.diff <- diff(y)

# ACF
Acf(y.diff)
# PACF
Acf(y.diff,type = "partial")

# plotting it...

par(mfrow = c(1,3))
# description of the data
plot(t[-1],y.diff)
Acf(y.diff)
# PACF
Acf(y.diff,type = "partial")
par(mfrow = c(1,1))

```

Based on the first difference, we observe that the rates are more stationary over time. Also, based on the Partial ACF, we can see that it seems that there are a order three of dependency of previous values, which would be associated with a MA(3) model.

To be sure about the stationarity of the model considering the first difference, let's calculate the KPSS test for stationarity and seasonality.

```{r}
# tests
tseries::kpss.test(y)
tseries::kpss.test(y.diff)

```

So, we conclude that the first order difference would be a better option to guarantee the stationarity of this serie.

```{r}
## modeling it...

# RWD

mod.RWD <- Arima(
  y = y,
  order = c(0,1,0),
  include.drift = TRUE
)

# AR

mod.ar <- auto.arima(y)

summary(mod.RWD)
summary(mod.ar)

# comparing models
modelsummary(list(mod.RWD,mod.ar),output = "markdown")

## comparing forecasts
t_ends = 2022
tF = t_ends - max(t)

mF.RWD <- forecast(mod.RWD, h = tF) |> 
  as_tibble() |> 
  mutate(year = seq(max(t)+1,t_ends,1))

mF.AR <- forecast(mod.ar, h = tF) |> 
  as_tibble() |> 
  mutate(year = seq(max(t)+1,t_ends,1))

```

After modelling and comparing these models, we can see that adding three lagged observations to model the next one -- ARIMA (0,1,3) -- not make the model more parsimonious than modeling it as a Random Walk with Drift (RWD) considering the first difference.

Now, let's plot our observed and fitted (predicted and forecasted values).

```{r}

# creating a dataframe with all the estimates together

mF <- mF.RWD |> 
  mutate(type = "RWD") |> 
  bind_rows(
    mF.AR |> 
  mutate(type = "AR")
  )


ggplot() +
  geom_line(
    data = mort_m20,
    aes(x = Year, y = logRates),
    linewidth = 1.2,
    color = "black"
  ) +
  geom_line(
    data = mF,
    aes(x = year, y = `Point Forecast`, color = type),
    linewidth = 1.2
  ) +
  geom_ribbon(
    data = mF,
    aes(
      x = year,
      ymin = `Lo 95`, ymax = `Hi 95`,
      fill = type
    ), 
    alpha = 0.2
  ) +
  theme_light(base_size = 12)

```

[write the interpretation here...]
