---
title: "EDSD - Data Quality - IPUMS Exercise"
author: "Thiago Cordeiro Almeida"
format: 
  pdf:
    documentclass: article
    keep-tex: true
    include-in-header: |
      \usepackage[a4paper, left=2cm, right=2cm, top=2.5cm, bottom=2.5cm]{geometry}
      \usepackage{ragged2e}
      \justifying
      \sloppy
editor_options: 
  chunk_output_type: console
fontsize: 12pt
dpi: 150

---

```{r importing data,echo=FALSE,warning=FALSE,message=FALSE}

# Settings
rm(list = ls()) # removing all itens of environment
invisible(gc()) # clealing the storage memory of the computer

# Libraries ---------------------------------------------------------------

library(pacman)
p_load(tidyverse, ipumsr)


# Importing data ----------------------------------------------------------

## Let's use ipums package to import it

# First - importing ddi

ddi <- read_ipums_ddi("../data/ipumsi_00036.xml")
data <- read_ipums_micro(ddi)

```
# IPUMS International Data Extract and Analysis

Data qulity, European Doctoral School of Demography 2024-2025

INED-Paris (France)

Instructor: Mariona Lozano, Centre d'Estudis Demogràfics

Student: Thiago Cordeiro Almeida

\-\-\-

# Question and Variables
In this exercise you will gain basic familiarity with the IPUMS International data exploration and extract system to answer the following question: “What are the differences in urbanization, literacy, and occupational participation in Uganda and Mexico?”

You will create a data extract that includes the following variables: URBAN, SEX, EMPSTAT, OCCISCO, LIT, AGE. 

\-\-\-

# Part 1. Variable documentation

For each variable below, search through the tabbed sections of the variable description to answer each question.

  1. Under the “Household” dropdown menu, find the “Geography” subcategory and click on the variable URBAN. What constitutes an urban area in each country?
  
    a. Mexico 2000:
    
    Urban places are defined consistently across Mexican samples as
    localities with 2,500 or more persons.
    
    b. Uganda 2002:
    
    Urban areas in 2002 and 2014 are gazetted cities, municipalities
    and towns with more than 2,000 inhabitants.
  
  2. What are the codes for URBAN?
  
  `1` for Rural, `2` for Urban and `9` for Unknown.
  
  3. Find the variable EMPSTAT. Is the reference period of work the same for Mexico and Uganda? 
  
  As IPUMS documentation describes, the reference period of work can be different depending on the country. In this case, for Mexico and Uganda the reference period of work are the same, a "reference week (typically the week prior to the census)".
  
  4. What is the universe for EMPSTAT in:
  
    a. Mexico 2000: Persons age 12+
    b. Uganda 2002: Persons age 5+
    
# Part 2. Frequencies

  5. Find codes page for the SAMPLE variable. What are the codes for:
    
    a. Mexico 2000: 484200001
    b. Uganda 2002: 800200201
    
  6. How many individuals are in the Mexico 2000 sample extract?
    
    There are `r paste0(as.numeric(table(data$COUNTRY)[1]))` individuals.
    
  7. How many individuals are in the Uganda 2002 sample extract?
    
    There are `r paste0(as.numeric(table(data$COUNTRY)[2]))` individuals.
  
  8. What proportion of individuals in the sample lived in urban areas in each country?
  
    a. Mexico 2000: `r paste0(round(100*prop.table(table(data$URBAN, data$COUNTRY),2),2)[2,1])`
    b. Uganda 2002: `r paste0(round(100*prop.table(table(data$URBAN, data$COUNTRY),2),2)[2,2])`
    
# Part 3. Weighted frequencies

To get a more accurate estimate of the actual proportion of individuals living in urban areas, you will have to use the person weight.

  9. Using weights, what is the total population of each country?
  
    a. Mexico 2000: `r paste0(data |> summarise(n = sum(PERWT), .by = COUNTRY) |> round(0) |> pluck(2,1))`
    b. Uganda 2002: `r paste0(data |> summarise(n = sum(PERWT), .by = COUNTRY) |> round(0) |> pluck(2,2))`
  
  10. Using weights, how many individuals lived in urban areas in each country?
  
    a. Mexico 2000: `r paste0(data |> filter(URBAN == 2) |>  summarise(n = sum(PERWT), .by = COUNTRY) |> round(0) |> pluck(2,1))`
    b. Uganda 2002: `r paste0(data |> filter(URBAN == 2) |> summarise(n = sum(PERWT), .by = COUNTRY) |> round(0) |> pluck(2,2))`
    
  11. Using weights, what proportion of individuals lived in urban areas in each country?
  
    a. Mexico 2000: `r paste0(data |> summarise(n = sum(PERWT), .by = c(COUNTRY, URBAN)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> filter(URBAN == 2) |> pluck(4,1))`% of individuals lived in urban areas.
    b. Uganda 2002: `r paste0(data |> summarise(n = sum(PERWT), .by = c(COUNTRY, URBAN)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> filter(URBAN == 2) |> pluck(4,2))`% of individuals lived in urban areas.

# Part 4. Trends

  12. Using weights, which occupational category has the highest percentage of workers?
  
    a. Mexico 2000: `r paste0(ipums_val_labels(data$OCCISCO) |> filter(val == data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> arrange(COUNTRY, desc(prop)) |> mutate(ranking = row_number(), .by = COUNTRY) |>  filter(ranking == 1) |> pluck(2,1)) |> pluck(2,1)," was the highest\npercentage category of workers,"," with ",data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> arrange(COUNTRY, desc(prop)) |> mutate(ranking = row_number(), .by = COUNTRY) |>  filter(ranking == 1) |> pluck(4,1))`% of workers.
    
    b. Uganda 2002: `r paste0(ipums_val_labels(data$OCCISCO) |> filter(val == data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> arrange(COUNTRY, desc(prop)) |> mutate(ranking = row_number(), .by = COUNTRY) |>  filter(ranking == 1) |> pluck(2,2)) |> pluck(2,1)," was the\nhighest percentage category of workers,"," with ",data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO)) |> mutate(prop = round(n/sum(n)*100,2), .by = COUNTRY) |> arrange(COUNTRY, desc(prop)) |> mutate(ranking = row_number(), .by = COUNTRY) |>  filter(ranking == 1) |> pluck(4,2))`% of workers.
    
  13. Which occupation category has the highest percentage of female workers?
  
    a. Mexico 2000: `r paste0(ipums_val_labels(data$OCCISCO) |> filter(val == data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, SEX)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,OCCISCO)) |> arrange(COUNTRY, SEX, desc(prop)) |> mutate(ranking = row_number(), .by = c(COUNTRY,SEX)) |>  filter(SEX == 2, ranking == 1) |> pluck(2,1)) |> pluck(2,1)," was the highest percentage category\nof female workers,"," with ",data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, SEX)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,OCCISCO)) |> arrange(COUNTRY, SEX, desc(prop)) |> mutate(ranking = row_number(), .by = c(COUNTRY,SEX)) |>  filter(SEX == 2, ranking == 1) |> pluck(5,1))`% of female workers.
    
    b. Uganda 2002: `r paste0(ipums_val_labels(data$OCCISCO) |> filter(val == data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, SEX)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,OCCISCO)) |> arrange(COUNTRY, SEX, desc(prop)) |> mutate(ranking = row_number(), .by = c(COUNTRY,SEX)) |>  filter(SEX == 2, ranking == 1) |> pluck(2,2)) |> pluck(2,1)," was the highest percentage category\nof female workers,"," with ",data |> filter(OCCISCO != 99) |> summarise(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, SEX)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,OCCISCO)) |> arrange(COUNTRY, SEX, desc(prop)) |> mutate(ranking = row_number(), .by = c(COUNTRY,SEX)) |>  filter(SEX == 2, ranking == 1) |> pluck(5,2))`% of female workers.
    
  14. What is the labour force participation distribution by gender in each country?
  
    a. Mexico 2000: `r paste0("Male's labour force participation was ", data |> filter(EMPSTAT != 0) |>  mutate(in_participate = case_when(EMPSTAT %in% 1:2 ~ 1, TRUE ~ 0)) |> summarise(n = sum(PERWT), .by = c(COUNTRY, SEX,in_participate)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,SEX)) |> filter(in_participate == 1) |> arrange(COUNTRY, SEX, desc(prop)) |> pluck(5,1), "%.\nFemale's labour force participation was ",data |> filter(EMPSTAT != 0) |>  mutate(in_participate = case_when(EMPSTAT %in% 1:2 ~ 1, TRUE ~ 0)) |> summarise(n = sum(PERWT), .by = c(COUNTRY, SEX,in_participate)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,SEX)) |> filter(in_participate == 1) |> arrange(COUNTRY, SEX, desc(prop)) |> pluck(5,2),"%.")`.
    
    b. Uganda 2002: `r paste0("Male's labour force participation was ", data |> filter(EMPSTAT != 0) |>  mutate(in_participate = case_when(EMPSTAT %in% 1:2 ~ 1, TRUE ~ 0)) |> summarise(n = sum(PERWT), .by = c(COUNTRY, SEX,in_participate)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,SEX)) |> filter(in_participate == 1) |> arrange(COUNTRY, SEX, desc(prop)) |>  pluck(5,3), "%.\nFemale's labour force participation was ",data |> filter(EMPSTAT != 0) |>  mutate(in_participate = case_when(EMPSTAT %in% 1:2 ~ 1, TRUE ~ 0)) |> summarise(n = sum(PERWT), .by = c(COUNTRY, SEX,in_participate)) |> mutate(prop = round(n/sum(n)*100,2), .by = c(COUNTRY,SEX)) |> filter(in_participate == 1)  |> arrange(COUNTRY, SEX, desc(prop)) |> pluck(5,4),"%.")`.
    
  15. What percentage of women within the labour force is working:
  
    a. In agriculture in Mexico 2000: `r paste0(data |> filter(EMPSTAT %in% 1:2, SEX == 2) |> mutate(in_working = case_when(EMPSTAT == 1 ~ 1, TRUE ~ 0)) |> reframe(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, in_working)) |> reframe(working_rate = round(100*n[OCCISCO == 6]/sum(n),2),.by = COUNTRY) |> pluck(2,1),"% of women within the labour\nforce were working in agriculture.")`
    
    b. In agriculture in Uganda 2002: `r paste0(data |> filter(EMPSTAT %in% 1:2, SEX == 2) |> mutate(in_working = case_when(EMPSTAT == 1 ~ 1, TRUE ~ 0)) |> reframe(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, in_working)) |> reframe(working_rate = round(100*n[OCCISCO == 6]/sum(n),2),.by = COUNTRY) |> pluck(2,2),"% of women within the labour\nforce were working in agriculture.")`
    
    c. In service in Mexico 2000: `r paste0(data |> filter(EMPSTAT %in% 1:2, SEX == 2) |> mutate(in_working = case_when(EMPSTAT == 1 ~ 1, TRUE ~ 0)) |> reframe(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, in_working)) |> reframe(working_rate = round(100*n[OCCISCO == 5]/sum(n),2),.by = COUNTRY) |> pluck(2,1),"% of women within the labour\nforce were working in agriculture.")`
    
    d. In service in Uganda 2002: `r paste0(data |> filter(EMPSTAT %in% 1:2, SEX == 2) |> mutate(in_working = case_when(EMPSTAT == 1 ~ 1, TRUE ~ 0)) |> reframe(n = sum(PERWT), .by = c(COUNTRY, OCCISCO, in_working)) |> reframe(working_rate = round(100*n[OCCISCO == 5]/sum(n),2),.by = COUNTRY) |> pluck(2,2),"% of women within the labour force\nwere working in agriculture.")`
    
# Part 5. Graphical Analysis

  16. What percentage of the population is literate in each country?
  
    a. Mexico 2000: `r paste0(data |> filter(LIT %in% 1:2) |> summarise(n = sum(PERWT), .by = c(COUNTRY, LIT)) |> mutate(prop = round(100*n/sum(n),2), .by = COUNTRY) |> filter(LIT == 2) |> pluck(4,1))`% of Mexican population were literate in 2000.
    b. Uganda 2002: `r paste0(data |> filter(LIT %in% 1:2) |> summarise(n = sum(PERWT), .by = c(COUNTRY, LIT)) |> mutate(prop = round(100*n/sum(n),2), .by = COUNTRY) |> filter(LIT == 2) |> pluck(4,2))`% of Ugandan population were literate in 2002.
    
  17. (OPTIONAL) Create a graph to visualize differences in the percentage of literacy by AGE and SEX in both countries.

```{r,warning=FALSE,echo=FALSE,message=FALSE}

data |> 
  filter(LIT %in% 1:2, AGE < 100) |> 
  summarise(n = sum(PERWT), .by = c(COUNTRY, LIT, AGE)) |> 
  mutate(prop = round(100*n/sum(n),2), .by = c(COUNTRY,AGE)) |>
  mutate(
    COUNTRY = factor(COUNTRY, levels = c(484, 800), labels = c("Mexico","Uganda"))
  ) |> 
  filter(LIT == 2, AGE >= 10) |> 
  ggplot() +
  aes(x = AGE, y = prop, color = COUNTRY, group = COUNTRY) +
  geom_point(size = 2, alpha = .5, show.legend = FALSE) +
  geom_smooth(se = FALSE,linewidth = 1.3, show.legend = FALSE) +
  scale_color_manual(values = c("#5ab4ac","#998ec3")) +
  scale_y_continuous(breaks = seq(0,100,20)) +
  scale_x_continuous(breaks = seq(0,100,10)) +
  labs(
    x = "Age (years)",
    y = "Proportion of literate population (%)",
    caption = "Source: IPUMS International.",
    title = "Proportion of literate population by age (%) - Uganda (2002)\nand Mexico (2000)"
  ) +
  annotate(
    geom="text",
    x=80,
    y=80,
    label="Mexico (2000)",
    color="#5ab4ac",
    fontface =2,
    size = 6
  ) +
  annotate(
    geom="text",
    x=50,
    y=30,
    label="Uganda (2002)",
    color="#998ec3",
    fontface =2,
    size = 6
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(color = "grey44", face = "bold", hjust = .5),
    plot.caption = element_text(color = "grey44", face = "italic", hjust = 1),
    axis.title.x = element_text(color = "grey44", face = "bold", hjust = .5),
    axis.title.y = element_text(color = "grey44", face = "bold", hjust = .5),
    axis.text = element_text(color = "grey44")
  )

```

